---
title: 'Laying out figures with ggplot2'
date: '02-12-2024'
categories: ['Example', 'ggplot2']
description: 'This is a test post. In this post, I try out different functionalities in the hope of making a template.'
execute:
  # eval	# Evaluate the code chunk (if false, just shows the code in the output)
  # echo 	# Include the source code in output
  output: true # 	Include the results of executing the code in the output (true, false, or asis to indicate that the output is raw markdown and should not have any of Quarto’s standard enclosing markdown)
  message: false
  warning: false # Include warnings in the output
  error: false # Include errors in the output (note that this implies that errors executing code will not halt processing of the document).
  # include: # Catch all for preventing any output (code or results) from being included (e.g. include: false suppresses all output from the code block).
  # code-overflow: wrap
editor_options: 
  chunk_output_type: inline
knitr:
  opts_chunk: 
    fig.align: "center"
    fig-width: 8
    fig-asp: 0.75
    code-overflow: wrap
toc: true
---

# Setup

## Why?

::: ol
1.   Previously was using a combination of R, Powerpoint, and Latex to lay out figures.
2.   However, it was a very finicking process. If I'm going to be using Quarto/R for most things would be nice to do things all in one place.
3.   I really want to enable other students/academics to do things without reinventing the wheel.
:::

I like to use `pacman` (the package manager) because its function `p_load` acts like `library()` if the package is already installed and `install.packages()` if it isn't. You can also pass all your package names to it as a single vector. 

```{r setup, eval=FALSE}
#| collapse: true

install.packages("pacman", repos="https://cran.r-project.org", update = TRUE)

# Basic plotting packages
pacman::p_load("ggplot2", update = TRUE)

# All the packages we need for laying out plots
pacman::p_load("grid", "gridExtra", "gtable", "cowplot", "multipanelfigure", "egg", "patchwork", update = TRUE)

# Extra stuff
pacman::p_load("ggpubr", "ggplotify", "tidyverse", update = TRUE)
```

So these are the packages I want to compare. A lot of the inspiration for this work came from the `egg` vignettes ([here](https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html) and [here](https://cran.r-project.org/web/packages/egg/vignettes/Overview.html)), so make sure to check those out if you want further information. 

| Package | Version | Function(s) | Compatible with `ggsave()` | Alignment |
|------                 |------                                  |------                   |------ |------ |
| [r cookbook (`grid`)](http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/){.external target="_blank"}   | `r packageVersion("grid")`    | `multiplot`    | no    | no    |
| [`gridExtra`](https://cran.r-project.org){.external target="_blank"}   | `r packageVersion("gridExtra")`   | `grid.arrange`          | yes   | no    |
| [`gtable`](https://cran.r-project.org){.external target="_blank"}    | `r packageVersion("gtable")`    | `multi_panel_figure`    | yes   | yes   |
| [`cowplot`](https://cran.r-project.org){.external target="_blank"}   | `r packageVersion("cowplot")`   | `plot_grid`             | yes\* | yes\* |
| [`multipanelfigure`](https://cran.r-project.org){.external target="_blank"}    | `r packageVersion("multipanelfigure")` | `multi_panel_figure`    | yes   | yes   |
| [`egg`](https://cran.r-project.org/web/packages/egg/vignettes/Overview.html){.external target="_blank"}   | `r packageVersion("egg")`              | `ggarrange`             | yes   | yes   |
| [`patchwork`](https://cran.r-project.org){.external target="_blank"}   | `r packageVersion("patchwork")`        | `plot_layout`           | yes   | yes   |

::: {.p style="margin: 1em"}
 
:::

First I'm going to set up a dummy plot that will act as our tester going forward. I'm deliberately going to make it pretty complex to make sure I can demonstrate all the features we might want. I'll be using data from the Bureau of Meteorology that I downloaded and cleaned up. The data I used is available [here](resources/melbourne_weather_times.csv), and if you want to generate a similar dataset the code I used is [here](/resources/code-to-clean-weather-data.R).
```{r}
#| include: false
#| file: ../resources/code-to-clean-weather-data.r
```

```{r default plot}
p.temp.9am = 
  ggplot(data = subset(weather.times, Time == "Morning"), 
         aes(x = Date, y = Temp.C, color = WindDir, shape = Time)) +
  geom_point() + 
  scale_x_date(name = "Month", date_labels = "%b") +
  theme_classic() + theme(legend.position = "none")

p.temp.3pm = 
  ggplot(data = subset(weather.times, Time == "Morning"), 
         aes(x = Date, y = Temp.C, color = WindDir, shape = Time)) +
  geom_point() + 
  scale_x_date(name = "Month", date_labels = "%b") +
  theme_classic() + theme(legend.position = "none")
```

## Basics

Laying two plots out, side by side, is easy with basically any of these packages. The most basic way is probably with the confusingly named custom function `multiplot`, which is built in base packages and defined in [R Cookbook](http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/){.external target="_blank"}.
```{r multiplot function}
#| eval: TRUE
#| include: FALSE
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

```{r grid multiplot}
#| eval: TRUE
#| fig-asp: 0.5

multiplot(p.temp.9am, p.temp.3pm, cols = 2)
```

You can also achieve the same thing with functions already built into the packages `gridExtra`, `ggpubr` and `egg`, which are built on top of the basic `grid` package:
```{r text size}
#| include: FALSE

p.temp.9am = p.temp.9am + theme(text = element_text(size = 16))
p.temp.3pm = p.temp.3pm + theme(text = element_text(size = 16))
```


::: d-grid
::: {.row style="display:flex; width:100%;"}
::: col
#### gridExtra
```{r basic gridExtra}
#| fig-asp: 1.25

gridExtra::grid.arrange(
  p.temp.9am, p.temp.3pm, 
  nrow = 2)
```
:::
::: col
#### ggpubr
```{r basic ggpubr}
#| fig-asp: 1.25

ggpubr::ggarrange(
  p.temp.9am, p.temp.3pm, 
  nrow = 2)
```
:::
::: col
#### egg
```{r basic egg}
#| fig-asp: 1.25

egg::ggarrange(
  p.temp.9am, p.temp.3pm, 
  nrow = 2)
```
:::
:::
:::

However, have a look at the figures above. Doesn't it bother you that the axis titles, labels and tick-marks are repeated over multiple figures? It bothers me. 
It especially bothers me when I'm trying to squeeze a lot of information into one page. Repeated information isn't just redundant, it's actively detrimental to your audience's understanding of your data. Personally, if I was going to plot those two figures together in a paper, I would remove their common axis (y if they're side-by-side, x if they're stacked). 

## Adjustments for subfigures

The simplest way to do achieve this is to remove the offending elements from the plots themselves. I like to do this with theme environment shortcuts, which take advantage of `ggplot2`'s modular structure and can make your code a lot less verbose. For example:
```{r text size 2}
#| include: FALSE

p.temp.9am = p.temp.9am + theme(text = element_text(size = 12))
p.temp.3pm = p.temp.3pm + theme(text = element_text(size = 12))
```

```{r no axes}
#| eval: TRUE 
#| fig-height: 4

no.x = theme(axis.title.x = element_blank(), axis.text.x = element_blank())
no.y = theme(axis.title.y = element_blank(), axis.text.y = element_blank())
```

You can then plot your figures much more neatly, and you'll get exactly the same results using `gridExtra::grid.arrange` and `ggpubr::ggarrange`. However, you'll get a different result when you use `egg::ggarrange`, and in fact `egg::ggarrange` is slightly superior in this regard. It might be easier to see if we switch away from the actual plot and look at the "bones":

```{r boneplot}
lemme_see_dem_bones = function(plot){
  bonegrob = egg::expose_layout(plot, draw = F)
  boneplot = ggpubr::as_ggplot(bonegrob)
  return(boneplot)
}

# Functions ggpubr::as_ggplot() ggplotify::as.ggplot()
```

`expose_layout` allows you to see how the different plot elements are being parsed by whatever package you're using. 

::: d-grid
::: {.row style="display:flex; width:100%;"}
#### gridExtra
::: col
```{r basic gridExtra 2}
#| fig-asp: 1.25

gridExtra::grid.arrange(
  p.temp.9am + no.x, 
  p.temp.3pm, 
  nrow = 2)
```
:::

::: col
```{r basic gridExtra bones}
#| fig-asp: 1.25

gridExtra::grid.arrange(
  lemme_see_dem_bones(p.temp.9am + no.x), 
  lemme_see_dem_bones(p.temp.3pm), 
  nrow = 2)
```
:::

:::

::: {.row style="display:flex; width:100%;"}

#### ggpubr
::: col
```{r basic ggpubr 2}
#| fig-asp: 1.25

ggpubr::ggarrange(
  p.temp.9am + no.x, 
  p.temp.3pm, 
  nrow = 2)
```
:::

::: col
```{r basic ggpubr bones}
#| fig-asp: 1.25

ggpubr::ggarrange(
  lemme_see_dem_bones(p.temp.9am + no.x), 
  lemme_see_dem_bones(p.temp.3pm), 
  nrow = 2)
```
:::

:::

::: {.row style="display:flex; width:100%;"}
#### egg
::: col
```{r basic egg 2}
#| fig-asp: 1.25

egg::ggarrange(
  p.temp.9am + no.x, 
  p.temp.3pm, 
  nrow = 2)
```
:::

::: col
```{r basic egg bones}
#| eval: false
#| fig-asp: 1.25

eg = egg::ggarrange(
  p.temp.9am + no.x, 
  p.temp.3pm, 
  nrow = 2)

expose_layout(eg)
```
:::
:::
:::

You might have already noticed some problems with this approach. For one thing, it's pretty finicky and a bit of a time-trap. You're essentially manually removing bits of each plot to fit the image in your head, and before you know it you'll have spent 6 hours trying to get a figure to look right. 
For another, the sub-figures above are actually slightly different sizes. The axes titles take up part of the plot area allotted to each sub-figure, so removing them causes the rest of the plot area to expand to fill it. Not ideal.

Even worse, it's standard practice to label subfigures within the plot area. 
grid.arrange is a little too simple function, but the two `ggarrange` functions (in `egg` and `ggpubr`) will do it for you pretty easily. However, they are not flexible about where those labels are placed within the plot area, and this is especially obvious when illustrating with a 4-panel figure:
```{r text size 4}
#| include: FALSE

p.temp.9am = p.temp.9am + theme(text = element_text(size = 12))
p.temp.3pm = p.temp.3pm + theme(text = element_text(size = 12))
```

```{r 4 panels with labels}
#| fig-width: 10
#| fig-height: 8

# WH = c(4, 3)

# egg::ggarrange(
#   ggplotify::as.ggplot(
#     egg::set_panel_size(p.temp.9am + no.x, width = unit(WH[1], "in"), height = unit(WH[2], "in"))
#     ),
#   ggplotify::as.ggplot(
#     egg::set_panel_size(p.temp.9am + no.x + no.y, width = unit(WH[1], "in"), height = unit(WH[2], "in"))
#     ),
#   ggplotify::as.ggplot(
#     egg::set_panel_size(p.temp.3pm, width = unit(WH[1], "in"), height = unit(WH[2], "in"))
#     ),
#   ggplotify::as.ggplot(
#     egg::set_panel_size(p.temp.3pm + no.y, width = unit(WH[1], "in"), height = unit(WH[2], "in"))
#     ),
#   nrow = 2, ncol = 2, labels = c("A", "B", "C", "D"))

geg = egg::ggarrange(p.temp.9am + no.x, 
               p.temp.9am + no.x + no.y, 
               p.temp.3pm, 
               p.temp.3pm + no.y, 
               ncol = 2, nrow = 2,
               widths = c(1, 1), heights = c(1, 1), 
               labels = c("A", "B", "C", "D"),
               debug= T,
               label.args = list(gp = gpar(fontfamily = "sans", fontsize = 12, fontface = "bold"), x = unit(1, "null")), 
               draw = T)
```
Ew.


```{r gtable}
#| eval: FALSE

gtab = gtable_matrix(name = NULL,
                     grobs = matrix(list(grob.1, grob.1), nrow = 1), 
                     widths = unit(c(1, 1), "null"), 
                     heights = unit(1, "null"))

g1 = expose_layout(dummy.1) 

gtab = gtable_matrix(name = NULL, 
                     grobs = matrix(list(g1, g1), nrow = 1), 
                     widths = unit(c(1, 1), "null"), 
                     heights = unit(1, "null"))

grid.draw(gtab)
```


# grob.1 = ggplotGrob(p.temp.9am) # Default plot as a grob

## Create ggplot from grob

```{r}
#| eval: FALSE

ggplotify::as.ggplot(g1)

egg::ggarrange(ggplotify::as.ggplot(g1), ggplotify::as.ggplot(g1), nrow = 1)
```

## The `r 'egg'` package

```{r}
#| eval: FALSE

# https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html

layout_matrix = rbind(c(1, 2, 3), c(1, 4, 4))

egg::ggarrange(ggplot(), ggplot(), ggplot(), ggplot(), 
               # ncol = 3, nrow = 2, 
               widths = c(1, 1, 1, 2), heights = c(2, 1, 1, 1), 
               labels = c("A", "B", "C", "D"),
               label.args = list(gp = gpar(fontfamily = "sans", 
                                           fontsize = 12, fontface = "bold"), 
                                 x = unit(1.95, "cm"), vjust = 1, hjust = 1), 
               draw = T)

g1 = gtable_frame(ggplotGrob(ggplot()), width = unit(1, "null"), height = unit(2, "null"), debug = TRUE)
g2 = gtable_frame(ggplotGrob(ggplot()), width = unit(1, "null"), height = unit(1, "null"), debug = TRUE)
g3 = gtable_frame(ggplotGrob(ggplot()), width = unit(1, "null"), height = unit(1, "null"), debug = TRUE)
g4 = gtable_frame(ggplotGrob(ggplot()), width = unit(2, "null"), height = unit(1, "null"), debug = TRUE)

g23 = gtable_frame(gtable_cbind(gtable_frame(ggplotGrob(ggplot())), 
                                gtable_frame(ggplotGrob(ggplot()))), 
                   width = unit(2, "null"), height = unit(1, "null"))
g234 = gtable_frame(gtable_rbind(g23, 
                                 gtable_frame(ggplotGrob(ggplot()))), 
                    width = unit(2, "null"), height = unit(2, "null"))
gcomb = gtable_frame(gtable_cbind(gtable_frame(ggplotGrob(ggplot())),
                                  g234),
                     width = unit(3, "null"), height = unit(2, "null"))

grid.newpage()
grid.draw(gcomb)
```

